---
title: "Additional_usability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional_usability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cercospoRa)
```


```{r}
# install epiphytoolR package if it is not installed
if("epiphytoolR" %in% installed.packages()[,1] == FALSE){
  remotes::install_github("PaulMelloy/epiphytoolR", dependencies = TRUE)
}
library(epiphytoolR)
```

Weather station data can be used for the model, however the example uses weather 
data downloaded using the R package `nasapower`.  

```{r eval=FALSE, include=TRUE}
# install stationaRy package if it is not installed
if("nasapower" %in% installed.packages()[,1] == FALSE){
  install.packages("nasapower")
}
library(nasapower)

np <- get_power(
  community = "ag",
  pars = c(
    "RH2M", # Relative humidity at 2 meters
    "T2M", # Air temperature at 2 meters
    "WS2M", # wind speed at 2 meters,
    "WD2M",# wind direction at 2 meters,
    "PRECTOTCORR",
    "ALLSKY_SFC_SW_DWN", # short wave downward solar radiation at the earth surface
    "ALLSKY_SFC_LW_DWN", # long wave downward solar radiation at the earth surface
    "PS", # surface pressure
    "ALLSKY_SRF_ALB" # Surface albedo
  ),
  temporal_api = "hourly",
  lonlat = c(9.916,51.41866),
  dates = c("2018-02-01", "2022-12-08"),
  time_standard = "UTC"
)

# set as a data.table
setDT(np)

# give the weather data a station name
np$station <- "Belgium"
# provide a wind direction standard deviation
np$WDSD <- 20

# format weather
w_beau <- 
  format_weather(np,
                 time_zone = "UTC",
                 temp = "T2M",
                 rain = "PRECTOTCORR",
                 YYYY = "YEAR",
                 MM = "MO",
                 DD = "DY",
                 hh = "HR",
                 rh = "RH2M",
                 ws = "WS2M",
                 wd = "WD2M",
                 wd_sd = "WDSD",
                 station = "station",
                 lon = "LON", 
                 lat = "LAT"
                 )
```

## Calculate epidemic onset  

The following function calculates the earliest date from which a cercospora leaf
spot epidemic could commence. 
`start` and `end` provide the time window for which the model should run.
Canopy closure (`c_closure`) indicates when the model should start point for the 
model.


```{r eval=FALSE, include=TRUE}
calc_epidemic_onset(start = as.POSIXct("2022-04-25",tz = "UTC"),
                    end = as.POSIXct("2022-09-30",tz = "UTC"),
                    c_closure = as.POSIXct("2022-07-01",tz = "UTC"),
                    weather = w_beau)
```

We can find the epidemic start date by calling the `calc_epidemic_onset()` function.
This function is a wrapper for the `calc_DIV()` function which calculates the daily
infection values.

##  Calculate daily infection values  
`calc_DIV()` will do this for all the days in the weather data frame provided, starting at 
the first day, integrating over the hourly data.  

```{r eval=FALSE, include=TRUE}
divalues <- calc_DIV(dat = w_beau[times > as.POSIXct("2022-07-01")])

# we take the cumulative sum of the daily infection values to monitor the increasing
# risk of infection from cercospora
divalues[, DIV_sum := cumsum(DIV)]
divalues
```



