---
title: "Getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup and Requirements  

```{r setup}
library(cercospoRa)
library(data.table)
library(terra)
library(sf)
library(ggplot2)
```

The `cercospoRa` package estimates the daily infection values which tracks the 
progress towards a *Cercospora beticola* epidemic on sugar beet.
The package requires formatted hourly weather data in a format which the model 
recognises.

### Format weather station data  

Included in the package is a weather data set from a station adjacent to a sugar 
beet field trial.  

The weather data is first cleaned into to a format that the model can use without 
returning errors due to inconsistencies and `NAs`, which are common in weather data. 
This will require the user to think about how to replace the `NAs` or state default
values to complete the data set.  

```{r prep_wdata}
# classify to data.table
wthr <- data.table(weathr)

# Use POSIXct formatted time.
wthr[,Time := as.POSIXct(paste0(Datum, " ",Stunde,":00"),tz = "UTC")]

# Nominate Latitude and Longitude location of the weather station. 
# While not needed in cercospoRa some plant disease models will use location to 
#  decide the closest weather station to pull weather from
wthr[, c("lon","lat") := list(9.916,51.41866)]

# weather is hourly and will error if we don't specify a standard deviation of 
#  weather direction. This is intentional to force the user to decide how variable
#  the wind direction data could be.
wthr[, wd_std := 20]

# remove all data after September as it contains missing data
wthr <- wthr[Datum < as.POSIXct("2022-10-01")]

# set NA wind speed values to zero
wthr[is.na(WG200), WG200 := 0]
```

Now we can use the `format_weather()` function.  

```{r format_wdata}
wthr <- format_weather(wthr,
                         POSIXct_time = "Time",
                         time_zone = "UTC",
                         temp = "T200",
                         rain = "N100",
                         rh = "F200",
                         wd = "WR200",
                         ws = "WG200",
                         station = "Station",
                         lon = "lon",
                         lat = "lat",
                         wd_sd = "wd_std",
                         data_check = FALSE # this stops the function from checking for faults
                         )
# As the data is formatted closely enough for what is expected for the model. 
# We can elect to turn the data_check off so 
```


## Calculate epidemic onset  

The following function calculates the earliest date from which a cercospora leaf
spot epidemic could commence. 
`start` and `end` provide the time window for which the model should run.
Canopy closure (`c_closure`) indicates when the model should start point for the 
model.
This is due to an observation by Wolf canopy closure pr

This date is determined by the negative prognosis models to be the last day for 
which a epidemic is likely not to start.  

```{r}
cercospoRa::calc_epidemic_onset(start = as.POSIXct("2022-04-25",tz = "UTC"),
                    end = as.POSIXct("2022-09-30",tz = "UTC"),
                    c_closure = as.POSIXct("2022-07-01",tz = "UTC"),
                    weather = wthr)
```


## Estimate canopy closure  

Canopy closure is traditionally observed manually. 
The following methods permits canopy closure date estimation from remotely sensed
data.
UAV or satellite data can be used to produce georeferenced leaf area index (LAI)
rasters. 

```{r}
# Get file location of example rasters with LAI values
image_files <- list.files(system.file("extdata", "uav_img",package = "cercospoRa"),
                          pattern = "tif",
                          full.names = TRUE)

# Read in data and check for consistency  
epidemic_onset_param <-
  read_sb_growth_parameter(img_files = image_files,
                           img_dates = as.POSIXct(c("2022-06-14","2022-06-28"),
                                                  tz = "UTC"),
                           target_res = 10)

terra::plot(epidemic_onset_param)
```

The next function calculates the sugar beet growth rate. 

```{r}
param_rxt <- calc_r_x0(epidemic_onset_param,
                      min_r = 0.02,
                      max_r = 0.05,
                      k = 6)
```

Calculate canopy closure uses a logistic regression to estimate the canopy closure 
date for each raster cell.

```{r}
canopy_closure <- calc_c_closure(param_rxt,
                                 x1 = 1.3,
                                 k=6)
terra::plot(canopy_closure)
```

Finally we can use a specialised wrapper function to calculate the earliest possible
onset of a cercospora leaf spot epidemic when chemical intervention might be needed.

```{r}
epidemic_onset_map <- 
  calc_epidemic_onset_from_image(start =as.POSIXct("2022-04-25",tz = "UTC"),
                                 end = as.POSIXct("2022-09-30",tz = "UTC"),
                                 c_closure = canopy_closure,
                                 weather = wthr)
epidemic_onset_map
```

plot the heterogeneity of onset date

```{r}
terra::plot(epidemic_onset_map)
```

Convert numeric date back to POSIXct format
```{r}
onset_dates <- as.POSIXct(terra::values(epidemic_onset_map),
                          tz = "UTC",
                          origin = "1970-01-01")
```

```{r}
library(ggplot2)
s_date <- as.POSIXct(terra::values(epidemic_onset_map),
                     tz = "UTC",
                     origin = "1970-01-01")

# convert to a data.table
dtdate <- data.table(cell = seq_along(s_date))
dtdate$Dates <- s_date

# Summarise the range of epidemic onset dates
summary(s_date[is.na(s_date)== FALSE])


# Plot dates on boxplot
dtdate[is.na(Dates) == FALSE,] |>
ggplot2::ggplot(aes(y=Dates))+
  geom_boxplot()
```


### Random transects  

Lets randomise some transects in the field.  

100 leaves are taken for the disease assessment. 
If we randomly choose 50 meter transects, 2 leaves sampled per meter, we can 
estimate the possible variation in scouting in this field.  

Transects are usually taken at a 45 degree angle and avoid the edge.  

```{r}
n_transect <- 1000

#initialise 1000 points in the raster
rcoordinates <- 
      spatSample(epidemic_onset_map,n_transect,
             replace = TRUE,
             na.rm = TRUE,
             xy = TRUE)

# Add extra columns for the transect end coordinates
rcoordinates <- 
  cbind(rcoordinates,
        data.frame(lon = NA_real_,lat = NA_real_), e_date = NA_real_)

# Loop over each row and find a transect that does not leave the field
for(i in 1:n_transect){
  in_or_out <- TRUE
  
  while(in_or_out){
    # get the coordinate 50 meters from the randomly chosen starting point
    d1 <- geosphere::destPoint(rcoordinates[i,1:2],
                               sample(c(45,135,225,315),size = 1),
                               50)

  in_or_out <- 
    d1[1] < ext(epidemic_onset_map)[1] |
    d1[1] > ext(epidemic_onset_map)[2] |
    d1[2] < ext(epidemic_onset_map)[3] |
    d1[2] > ext(epidemic_onset_map)[4]
}
 rcoordinates[i,"lon"] <- d1[1]
 rcoordinates[i,"lat"] <- d1[2]
 
 dat1 <-rbind(rcoordinates[i,1:2],
              c(d1[1],d1[2]), make.row.names = FALSE)
 
 v1 <- vect(st_linestring(as.matrix(dat1)))

 rcoordinates[i,"e_date"] <- 
   extract(epidemic_onset_map,v1,fun = "mean",na.rm = TRUE)[,2] 
}

rcoordinates$e_date <- 
  as.POSIXct(rcoordinates$e_date,
           tz = "UTC",
           origin = "1970-01-01")

rcoordinates |>
ggplot2::ggplot(aes(y=e_date))+
  geom_boxplot()
```

