---
title: "Getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cercospoRa)
library(data.table)
```

The `cercospoRa` package estimates the proportional progress towards a 
*Cercospora beticola* epidemic on sugarbeet.
This requires formated hourly weather data that the model recognises.
The R package `epiphytoolR` contains functions which automatically checks the weather
data and returns it in the required hourly format.  

We will need to install epiphytoolR  

```{r}
# install stationaRy package if it is not installed
if("epiphytoolR" %in% installed.packages()[,1] == FALSE){
  remotes::install_github("PaulMelloy/epiphytoolR",ref = "main", dependencies = TRUE)
}
library(epiphytoolR)
```


Next lets download some weather data using the R package `nasapower`.  

```{r eval=FALSE, include=TRUE}
# install stationaRy package if it is not installed
if("nasapower" %in% installed.packages()[,1] == FALSE){
  install.packages("nasapower")
}
library(nasapower)

np <- get_power(
  community = "ag",
  pars = c(
    "RH2M", # Relative humidity at 2 meters
    "T2M", # Air temperature at 2 meters
    "WS2M", # wind speed at 2 meters,
    "WD2M",# wind direction at 2 meters,
    "PRECTOTCORR",
    "ALLSKY_SFC_SW_DWN", # short wave downward solar radiation at the earth surface
    "ALLSKY_SFC_LW_DWN", # long wave downward solar radiation at the earth surface
    "PS", # surface pressure
    "ALLSKY_SRF_ALB" # Surface albedo
  ),
  temporal_api = "hourly",
  lonlat = c(9.916,51.41866),
  dates = c("2018-02-01", "2022-12-08"),
  time_standard = "UTC"
)

# set as a data.table
setDT(np)

# give the weather data a station name
np$station <- "Belgium"
# provide a wind direction standard deviation
np$WDSD <- 20

# format weather
w_beau <- 
  format_weather(np,
                 time_zone = "UTC",
                 temp = "T2M",
                 rain = "PRECTOTCORR",
                 YYYY = "YEAR",
                 MM = "MO",
                 DD = "DY",
                 hh = "HR",
                 rh = "RH2M",
                 ws = "WS2M",
                 wd = "WD2M",
                 wd_sd = "WDSD",
                 station = "station",
                 lon = "LON", 
                 lat = "LAT"
                 )
```

```{r}
# calc_epidemic_onset(start = as.POSIXct("2022-04-25",tz = "UTC"),
#                     end = as.POSIXct("2022-09-30",tz = "UTC"),
#                     c_closure = as.POSIXct("2022-07-01",tz = "UTC"),
#                     weather = w_beau)
```

### Format weather station data  

This is data recorded by a weather station and contributed by Facundo and is from 
a field trial.  

We will format the data so the model functions can recognise the data and won't 
return errors due to inconsistencies and NAs which are common in weather data. 
This will require the user to think about how to replace the NAs or state default
values to complete the dataset.  

```{r}
# format to data.table
wthr <- data.table(weathr)

# Use POSIXct formatted time.
wthr[,Time := as.POSIXct(paste0(Datum, " ",Stunde,":00"),tz = "UTC")]

# Nominate Latitude and Longitude location of the weather station. 
# While not needed in cercospoRa some plant disease models will use location to 
#  decide the closest weather station to pull weather from
wthr[, c("lon","lat") := list(9.916,51.41866)]

# weather is hourly and will error if we don't specify a standard deviation of 
#  weather direction. This is intentional to force the user to decide how variable
#  the wind direction data could be.
wthr[, wd_std := 20]

# remove all data after September as it contains missing data
wthr <- wthr[Datum < as.POSIXct("2022-10-01")]

# set NA wind speed values to zero
wthr[is.na(WG200),WG200 := 0]

# set NA wind direction values to 20 degrees. Wind is not important for this model
wthr[is.na(WR200),WR200 := 20]

wthr <- format_weather(wthr,
                         POSIXct_time = "Time",
                         time_zone = "UTC",
                         temp = "T200",
                         rain = "N100",
                         rh = "F200",
                         wd = "WR200",
                         ws = "WG200",
                         station = "Station",
                         lon = "lon",
                         lat = "lat",
                         wd_sd = "wd_std",
                         data_check = FALSE # this stops the function from checking for faults
                         )
# As the data is formatted closely enough for what is expected for the model. 
# We can elect to turn the data_check off so 
```

We can find the epidemic start date by calling the `calc_epidemic_onset()` function.
This function is a wrapper for the `calc_DIV()` function which calculates the daily
infection values for both, Wolf and Racca methods.

This date is determined by the negative prognosis models to be the last day for 
which a epidemic is likely not to start.  

```{r}
cercospoRa::calc_epidemic_onset(start = as.POSIXct("2022-04-25",tz = "UTC"),
                    end = as.POSIXct("2022-09-30",tz = "UTC"),
                    c_closure = as.POSIXct("2022-07-01",tz = "UTC"),
                    weather = wthr)
```

  
`calc_DIV()` will do this for all the days in the weather data frame provided, starting at 
the first day, integrating over the hourly data.  

```{r}
divalues <- calc_DIV(dat = wthr[times > as.POSIXct("2022-07-01")])

# we take the cumulative sum of the daily infection values to monitor the increasing
# risk of infection from cercospora
divalues[, DIV_sum := cumsum(DIV)]
divalues[, DIV_sum_racca := cumsum(DIV_racca)]
divalues
```


